---
title: "Unsupervised learning: principal component analysis"
author: "<br>SURE 2025<br><br>Department of Statistics & Data Science<br>Carnegie Mellon University"
footer:  "[36-SURE.github.io/2025](https://36-sure.github.io/2025)"
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.25
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 9
)
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```


# Background

## Motivating example: Starbucks drinks

```{r}
library(tidyverse)
theme_set(theme_light())
starbucks <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-12-21/starbucks.csv"
) |>
  # convert columns to numeric that were saved as character
  mutate(trans_fat_g = as.numeric(trans_fat_g), fiber_g = as.numeric(fiber_g))
glimpse(starbucks)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

```{r}
#| eval: false
starbucks_feat <- starbucks |> 
  select(serv_size_m_l:caffeine_mg)
starbucks_pca <- prcomp(starbucks_feat, center = TRUE, scale. = TRUE)
summary(starbucks_pca)
```


## Unsupervised learning

* No response variable (i.e., data are not labeled)

. . .

* Only given a set of features measured on a set of observations

. . .

* Unsupervised learning is more subjective than supervised learning (difficult to tell how "good" you are doing)

. . .

* There is no simple goal for the analysis, such as prediction of a response in supervised learning

. . .

* Unsupervised learning can be useful as a pre-processing step for supervised learning

. . .

Think of unsupervised learning as **an extension of EDA---there’s no unique right answer!**

## Fundamental problems in unsupervised learning

. . .

1. **Dimension reduction:** reduce the original feature dimension of the data to something smaller so we can explore/visualize the data

  * Methods: **PCA (this lecture)**, ICA, t-SNE, UMAP,...
  
. . .

2. **Clustering**: group the observations in the data into different clusters

  * Methods: hard clustering ($k$-means, hierarchical clustering,...), soft clustering (mixture model)
  
. . .
  
3. **Density estimation**: estimate the underlying distribution of the data (in a way that makes sense)

## Dimension reduction: the big picture

**Key question: How do we visualize the structure of high-dimensional data?**

. . .

Example: What if you're given a dataset with 50 variables and are asked to make one visualization that best represents the data? What do you do?

. . .

Tedious task: Make a series of all $\displaystyle \binom{50}{2} = 1225$ pairs of plots? Or make a giant correlation heatmap?

. . .

**Intuition: Take high-dimensional data and represent it in 2-3 dimensions, then visualize those dimensions**


## Dimension reduction: the big picture

*   It's usually really hard to visualize many dimensions at the same time

. . .

*   Often, it makes a lot of sense to choose 2-3 of the "most important dimensions" and just plot those

. . .

*   PCA is a very common way to define "most important dimensions"

. . .

*   PCA provides the linear combinations of variables that capture the most variation in the data

. . .

*   It's common to plot the first two principal components in a scatterplot

. . .

*   It's very useful to plot principal components with a biplot

    *   Adds interpretability to the principal components, and helps reveal relationships among the variables

## What is the goal of dimension reduction?

Given a dataset with $p$ variables (columns) for $n$ observations (rows), can we find a smaller number of dimensions that captures the __interesting__ structure in the data?

. . .

__Dimension reduction__: 

- Focus on reducing the dimensionality of the feature space (i.e., number of columns) 

- __Retain__ most of the information / __variability__ in a lower dimensional space (i.e., reducing the number of columns)




## [Principal components analysis (PCA)](https://en.wikipedia.org/wiki/Principal_component_analysis)

* PCA produces a low-dimensional representation of a dataset

* PCA finds a set of variables that are linear combinations of the features that have maximal variance *and* are mutually uncorrelated

* The set of uncorrelated variables is called the **principal components (PCs)**

## Principal components

* PCs are weighted, normalized linear combinations of the original variables

* Weights reveal how different variables are ___loaded___ into the PCs

* First PC accounts for most variation in the data, second PC for second-most variation, and so on


## Principal components (a little more formal)

* Assume $\boldsymbol{X}$ is a $n \times p$ matrix that is __centered__ and __stardardized__

. . .

* PCA will give us $p$ principal components that are $n$-length columns, denoted by $Z_1, \dots, Z_p$

  * The first PC is the linear combination that has the largest possible variance

  * Each succeeding component has the largest possible variance under the constraint that it is uncorrelated with the preceding components

  * **A small number of principal components often explains a high proportion of the variability in the data**

## First principal component

$$Z_1 = \phi_{11} X_1 + \phi_{21} X_2 + \dots + \phi_{p1} X_p$$

. . .

* $\phi_{11}, \phi_{21}, \dots, \phi_{p1}$ are the loadings of the first PC 

* These are weights indicating the contributions of each feature
  
* Weights are normalized $\displaystyle \sum_{j=1}^p \phi_{j1}^2 = 1$ (else, arbitraily large variance)
  
* Together, the loadings make up the __loading vector__ for the first PC, $\phi_{1} = (\phi_{11}, \phi_{21}, \dots, \phi_{p1})$

* $Z_1$ is a linear combination of the $p$ variables that has the __largest variance__ (among all PCs)


## Second principal component

$$Z_2 = \phi_{12} X_1 + \phi_{22} X_2 + \dots + \phi_{p2} X_p$$

* $\phi_{12}, \phi_{22}, \dots, \phi_{p2}$ are the loadings of the second PC 
  
* Weights are normalized $\displaystyle \sum_{j=1}^p \phi_{j2}^2 = 1$
  
* Together, the loadings make up the __loading vector__ for the second PC, $\phi_{2} = (\phi_{12}, \phi_{22}, \dots, \phi_{p2})$

* $Z_2$ is a linear combination of the $p$ variables that has the __largest variance__ *but* __subject to constraint it is uncorrelated with $Z_1$__ 
    
## Obtaining the principal components

Repeat the same process to create $p$ principal components

* __Uncorrelated__: Each ($Z_j, Z_{j'}$) is uncorrelated with each other

* __Ordered Variance__: $\text{Var}(Z_1) > \text{Var}(Z_2) > \dots > \text{Var}(Z_p)$

* __Total Variance__: $\displaystyle \sum_{j=1}^p \text{Var}(Z_j) = p$

PCA can be

* formulated as an optimization problem

* solved via linear algebra techniques (e.g. eigen decomposition or singular value decomposition (SVD))

## Geometry of PCA

* The loading vector $\phi_1$ with elements $\phi_{11}, \phi_{21}, \dots, \phi_{p1}$ defines a direction in feature space along which the data vary the most

* If we project the observations in our dataset onto this direction, the projected values are called the principal component scores

## Geometry of PCA in 2D

```{r, echo = FALSE, fig.align='center', fig.height=7}
set.seed(123) 
plot_data <- tibble(x = rnorm(50, mean = 100, sd = 20)) |>
  mutate(y =  0.8 * x + rnorm(50, mean = 0, sd = 10))
basic_scatter <- ggplot(plot_data) +
  geom_point(aes(x, y), color = "black")+
  coord_equal()
basic_scatter
```

## Geometry of PCA in 2D

```{r, echo = FALSE, fig.align='center', fig.height=7}
#fit the model
line1 <- lm(y ~ x, plot_data)$coef
#extract the slope from the fitted model
line1.slope <- line1[2]
#extract the intercept from the fitted model
line1.intercept <- line1[1]
basic_scatter_yfit <- basic_scatter +
  geom_abline(aes(slope = line1.slope, intercept = line1.intercept),
              colour = "darkorange", linewidth = 1.5) +
  annotate("text", x = 75, y = 115, label = "y ~ x", color = "darkorange",
           size = 11)
basic_scatter_yfit
```


```{r, echo = FALSE}
#fit the model
line2 <- lm(x ~ y, plot_data)$coef
#extract the slope from the fitted model
line2.slope <- 1 / line2[2]
#extract the intercept from the fitted model
line2.intercept <- -(line2[1] / line2[2])
basic_scatter_xyfit <- basic_scatter_yfit +
  geom_abline(aes(slope = line2.slope, intercept = line2.intercept),
              colour = "blue", linewidth = 1.5) +
  annotate("text", x = 125, y = 55, label = "x ~ y", color = "blue",
           size = 9)
```

## Geometric interpretation of PCA in 2D

```{r, echo = FALSE, fig.align='center', fig.height=7}
pca <- prcomp(cbind(plot_data$x, plot_data$y))$rotation
pca.slope <- pca[2,1] / pca[1,1]
pca.intercept <- mean(plot_data$y) - (pca.slope * mean(plot_data$x))

basic_scatter_y_pca_fit <- basic_scatter_yfit +
  geom_abline(aes(slope = pca.slope, intercept = pca.intercept),
              colour = "blue", linewidth = 1.5) +
  annotate("text", x = 125, y = 55, label = "PCA", color = "blue",
           size = 11)
basic_scatter_y_pca_fit
```

## [Visualizing PCA](https://www.stevejburr.com/post/scatter-plots-and-best-fit-lines/) in two dimensions

Key idea: provide low-dimensional linear surfaces that are closest to the observations

```{r fig.width=10, fig.height=5, echo = FALSE, fig.align='center'}
plot_data |>
  #calculate the positions using the line equations:
  mutate(yhat_line1=(x*line1.slope+line1.intercept),
         xhat_line1=x,
         yhat_line2=y,
         xhat_line2=(y-line2.intercept)/line2.slope,
         #https://en.wikipedia.org/wiki/Distance_from_a_point_to_a_line
         a=pca.slope,
         b=-1,
         c=pca.intercept,
         xhat_line3=(b*(b*x-a*y)-(a*c))/((a*a)+(b*b)),
         yhat_line3=(a*(-b*x+a*y)-(b*c))/((a*a)+(b*b)),
         #add the slopes/intercepts to this data frame:
         slope_line1=line1.slope,
         slope_line2=line2.slope,
         slope_line3=pca.slope,
         intercept_line1=line1.intercept,
         intercept_line2=line2.intercept,
         intercept_line3=pca.intercept
         )|> 
  #drop intermediate variables
  select(-c(a,b,c)) |>
  pivot_longer(yhat_line1:intercept_line3, names_to = "key", values_to = "value") |>
  #transpose to a long form
  #gather(key="key",value="value",-c(x,y)) |> 
  # have "yhat_line1", want two colums of "yhat" "line1"
  separate(key,c("type", "line"), "_") |> 
  #then transpose to be fatter, so we have cols for xhat, yhat etc
  pivot_wider(names_from = "type", values_from = "value") |>
  #spread(key="type",value="value") |>
  #relable the lines with more description names, and order the factor for plotting:
  mutate(line=case_when(
           line=="line1" ~ "y ~ x",
           line=="line2" ~ "x ~ y",
           TRUE ~ "PCA"
         ),
         line = fct_relevel(line, "y ~ x", "x ~ y", "PCA")) |> 
  filter(line != "x ~ y") |> 
  ggplot() +
  geom_point(aes(x = x, y = y, color = line)) +
  geom_abline(aes(slope = slope, intercept = intercept, color = line), linewidth = 1.2) +
  geom_segment(aes(x = x, y = y, xend = xhat, yend = yhat, color = line)) +
  facet_wrap(~ line, ncol = 3) +
  scale_color_manual(values = c("darkorange", "blue")) +
  theme_bw() +
  theme(strip.background = element_blank(),
        legend.position = "none",
        strip.text = element_text(size = 16))
```

. . .

* The above is the **minimizing projection residuals** viewpoint of PCA

* There's another viewpoint---**maximizing variance**. That is, if we project all points onto the solid blue line, we maximize the variance of the resulting projected points across all such blue lines


# Examples

## Data: nutritional information of Starbucks drinks

```{r}
library(tidyverse)
theme_set(theme_light())
starbucks <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-12-21/starbucks.csv"
) |>
  # convert columns to numeric that were saved as character
  mutate(trans_fat_g = as.numeric(trans_fat_g), fiber_g = as.numeric(fiber_g))
glimpse(starbucks)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```


## Implementing PCA

Use the `prcomp()` function (based on SVD) for PCA on centered and scaled data

```{r}
starbucks_feat <- starbucks |> 
  select(serv_size_m_l:caffeine_mg)
starbucks_pca <- prcomp(starbucks_feat, center = TRUE, scale. = TRUE)
summary(starbucks_pca)
```


## Computing principal components

Extract the matrix of principal components (dimension will match original data)

Note: columns are uncorrelated, such that $\text{Var}(Z_1) > \text{Var}(Z_2) > \dots > \text{Var}(Z_p)$ 

```{r}
starbucks_pc_matrix <- starbucks_pca$x
head(starbucks_pc_matrix)
```

## Extracting loading matrix

```{r}
starbucks_pca$rotation
```


## Visualizing first two principal components

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r pca-plot, eval = FALSE}
starbucks <- starbucks |> 
  mutate(pc1 = starbucks_pc_matrix[,1], 
         pc2 = starbucks_pc_matrix[,2])
starbucks |> 
  ggplot(aes(x = pc1, y = pc2)) +
  geom_point(alpha = 0.5) +
  labs(x = "PC 1", y = "PC 2")
```

* Principal components are not interpretable

* Make a __biplot__ with arrows showing the linear relationship between one variable and other variables

:::

::: {.column width="50%" style="text-align: left;"}

```{r ref.label="pca-plot", echo = FALSE, fig.height=9}
```

:::
:::


## Making PCs interpretable with biplots

Biplot displays both the space of observations and the space of variables

Check out the [`factoextra`](http://www.sthda.com/english/wiki/factoextra-r-package-easy-multivariate-data-analyses-and-elegant-visualization) package

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r pca-biplot, eval = FALSE}
library(factoextra)
# fviz_pca_var(): projection of variables
# fviz_pca_ind(): display observations with first two PCs
starbucks_pca |> 
  fviz_pca_biplot(label = "var",
                  alpha.ind = 0.25,
                  alpha.var = 0.75,
                  labelsize = 5,
                  col.var = "darkblue",
                  repel = TRUE)
```

Interpreting a biplot

* Arrow direction

* Arrow angles
  
* Arrow length

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label = "pca-biplot", echo = FALSE, fig.height=10}

```

:::
:::

## Interpreting a biplot

::: columns
::: {.column width="50%" style="text-align: left;"}

Interpreting the **direction of a particular arrow**

* For example, the far left arrow for `caffeine_mg` suggests that, as `caffeine_mg` increases, $Z_1$ and $Z_2$ tend to decrease

  * in other words, within the definition of $Z_1$ and $Z_2$, the coefficient for `caffeine_mg` is negative

* In contrast, the arrow for `serv_size_m_l` is pointing to the upper right, indicating that as `serv_size_m_l` increases, both $Z_1$ and $Z_2$ also increase

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label = "pca-biplot", echo = FALSE, fig.height=10}

```

:::
:::

## Interpreting a biplot

::: columns
::: {.column width="50%" style="text-align: left;"}

**The angle of between the vectors is indicative of the correlation between different variables**

* $90^{\circ}$ (right angle): uncorrelated (e.g. `serv_size_m_l` and `saturated_fat_g`)

* $< 90^{\circ}$ (vectors in similar directions):<br>positively correlated<br>(e.g. `sugar_g` and `total_carbs_g`)

* $> 90^{\circ}$ (vectors in different directions):<br>negatively correlated<br>(e.g. `caffeine_mg` and `calories`)

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label = "pca-biplot", echo = FALSE, fig.height=10}

```

:::
:::

## Interpreting a biplot

::: columns
::: {.column width="50%" style="text-align: left;"}

**The arrow length indicates how strongly related the principal components are with the individual variables**

* For example, `serv_size_m_l` has a fairly long arrow because it has a large positive coefficient in the rotation matrix (see below)

* Meanwhile, `caffeine_mg` has a relatively short arrow because its coefficients are relatively small.

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label = "pca-biplot", echo = FALSE, fig.height=10}

```

:::
:::

## Intuition on how many PCs to use

```{r}
summary(starbucks_pca)
```

. . .

* PC1 accounts for the most variation in our data, PC2 accounts for the next most, and so on

. . .

* Each time we add a new PC dimension, we capture a higher proportion of the variability/information in the data

. . .

* But that increase in proportion decreases for each new dimension we add (i.e., additional PCs will add smaller and smaller variance)

. . .

* **It is recommended to keep adding PCs until the marginal gain levels off (i.e. decreases to the point that it isn’t too beneficial to add another dimension to the data)**



## Create a scree plot (or elbow plot)

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r, eval=FALSE, fig.align='center', out.width="90%"}
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(
    yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
    linetype = "dashed", 
    color = "darkred",
  )
```

```{r, eval = TRUE, echo = FALSE, fig.align='center', out.width="80%"}
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(
    yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
    linetype = "dashed", 
    color = "darkred",
  ) +
  theme(text = element_text(size = 20))
```


:::

::: {.column width="50%" style="text-align: left;"}

* Shows proportion of variation explained by each PC

* Visual rule-of-thumb: look for the "elbow" (where the proportion of variation starts to become flat)

* A heuristic rule of thumb: horizontal line at $1/p$

* Based on this plot, we can make a strong argument to use first 3 PCs (but maybe go up to first 5 PCs for another substantial drop in the elbow)



:::
:::

## Discussion

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r, eval=FALSE, fig.align='center', out.width="90%"}
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(
    yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
    linetype = "dashed", 
    color = "darkred",
  )
```

```{r, eval = TRUE, echo = FALSE, fig.align='center', out.width="80%"}
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(
    yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
    linetype = "dashed", 
    color = "darkred",
  ) +
  theme(text = element_text(size = 20))
```


:::

::: {.column width="50%" style="text-align: left;"}

* Let’s say we decide to use first 3 PCs (to capture satisfactory amount of the variation in the data)

* Our biplot only plots the first 2 PCs, and so we are hiding some data information that we are better off plotting in some way if possible

  * Specifically, we are hiding about 30% of the information (i.e. the total amount of information captured by PCs 3, 4, ..., 11)

  * About a third of this remaining information is captured by that third component that we are not plotting

* This means that, theoretically, we should plot three quantitative variables (e.g., point size, transparency, or even a 3D scatterplot)

* Alternatively, we could make three scatterplots (one for each pair of PCs)

:::
:::


## Remember the spelling...

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">folks. it&#39;s _principal_ (not principle) components analysis.</p>&mdash; Stephanie Hicks, PhD (@stephaniehicks) <a href="https://twitter.com/stephaniehicks/status/1626428875323367424?ref_src=twsrc%5Etfw">February 17, 2023</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

# Appendix

## PCA output

```{r}
# str(starbucks_pca)
```

Examine the output after running `prcomp()`

* `starbucks_pca$sdev`: singular values ($\sqrt{\lambda_j}$)

* `starbucks_pca$rotation`: loading matrix ($V$)

*  `starbucks_pca$x`: principal component scores matrix ($Z=XV$)

Can use the `broom` package for tidying `prcomp()`

```{r, eval=FALSE}
tidy(starbucks_pca, matrix = "eigenvalues") # equivalent to starbucks_pca$sdev
tidy(starbucks_pca, matrix = "rotation") # equivalent to starbucks_pca$rotation
tidy(starbucks_pca, matrix = "scores") # equivalent to starbucks_pca$x
```

## Proportion of variance explained

```{r}
library(broom)
starbucks_pca |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(x = PC, y = percent)) +
  geom_line() + 
  geom_point() +
  geom_hline(yintercept = 1 / ncol(starbucks_feat), color = "darkred", linetype = "dashed") +
  scale_x_continuous(breaks = 1:ncol(starbucks_pca$x))
```

## Cumulative proportion of variance explained

```{r}
library(broom)
starbucks_pca |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(x = PC, y = cumulative)) +
  geom_line() + 
  geom_point() +
  scale_x_continuous(breaks = 1:ncol(starbucks_pca$x))
```

## [__Singular value decomposition (SVD)__](https://en.wikipedia.org/wiki/Singular_value_decomposition)

$$
X = U D V^T
$$

- $U$ and $V$: matrices containing the left and right __singular vectors of scaled matrix $X$__

- $D$: diagonal matrix of the __singular values__

. . .

- SVD simplifies matrix-vector multiplication as __rotate, scale, and rotate again__

. . .

$V$: __loading matrix__ for $X$ with $\phi_{j}$ as columns

  - $Z = X  V$: PC matrix

. . .

Bonus: __Eigenvalue decomposition__ (or spectral decomposition)

- $V$: __eigenvectors__ of $X^TX$ (covariance matrix, $^T$: _transpose_)

- $U$: __eigenvectors__ of $XX^T$

- The singular values (diagonal of $D$) are square roots of the __eigenvalues__ of $X^TX$ or $XX^T$

- Meaning that $Z = UD$

## Eigenvalues guide dimension reduction

We want to choose $p^* < p$ such that we are explaining variation in the data

. . .

Eigenvalues $\lambda_j$ for $j \in 1, \dots, p$ indicate __the variance explained by each component__

  - $\displaystyle \sum_j^p \lambda_j = p$, meaning $\lambda_j \geq 1$ indicates $\text{PC}j$ contains at least one variable's worth in variability
  
  - $\displaystyle \frac{\lambda_j}{p}$: proportion of variance explained by $\text{PC}j$
  
  - Arranged in descending order so that $\lambda_1$ is largest eigenvalue and corresponds to PC1
  
. . .
  
  - Compute the cumulative proportion of variance explained (CVE) with $p^*$ components $$\text{CVE}_{p^*} = \sum_j^{p*} \frac{\lambda_j}{p}$$
Use [__scree plot__](https://en.wikipedia.org/wiki/Scree_plot) to plot eigenvalues and guide choice for $p^* <p$ by looking for "elbow" (rapid to slow change)
