---
title: "Supervised learning: linear regression"
author: "<br>SURE 2025<br><br>Department of Statistics & Data Science<br>Carnegie Mellon University"
footer:  "[36-SURE.github.io/2025](https://36-sure.github.io/2025)"
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.25
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 9
)
library(tidyverse)
theme_set(theme_light())
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

# Background

## Supervised learning

* General setup: Given a dataset of $n$ observations (rows) and $p$ variables (columns)

. . .

* Response/outcome/target/dependent variable $Y$ in one of the $p$ variables

. . .

* **Goal**: uncover associations between a set of predictors/features/covariates/explanatory variables and a response variable

  * Accurately predict unseen test cases

  * Understand which features affect the response (and how)

  * Assess the quality of our predictions and inferences
  
. . .  

* Two main types of problems

  * Regression: estimate average value of response (i.e. the response is quantitative)

  * Classification: determine the most likely class of a set of discrete response variable classes (i.e. the response is categorical)
  
## Examples of supervised learning methods

* Linear regression (today)

* Generalized linear models (GLMs) and penalized versions (lasso, ridge, elastic net)

* Smoothing splines, generalized additive models (GAMs)

* Decision trees and its variants (e.g., random forests, boosting)

* Deep learning, neural networks

## Resources

*   [Advanced Data Analysis from an Elementary Point of View](https://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/ADAfaEPoV.pdf) (aka the greatest book that will never be published)

*   [The Truth about Linear Regression](https://www.stat.cmu.edu/~cshalizi/TALR/TALR.pdf)

> "Regression", in statistical jargon, is the problem of guessing the average level of some quantitative response variable from various predictor variables. - [Note GOAT](http://bactra.org/notebooks/regression.html)

## Simple linear regression

**Linear regression is used when the response variable is quantitative**

. . .

Assume a __linear relationship__ for $Y = f(X)$ $$Y_{i}=\beta_{0}+\beta_{1} X_{i}+\epsilon_{i}, \quad \text { for } i=1,2, \dots, n$$

- $Y_i$ is the $i$th value for the __response__ variable
  
- $X_i$ is the $i$th value for the __predictor__ variable

. . .
  
- $\beta_0$ is an _unknown_, constant __intercept__

  -   average value for $Y$ if $X = 0$ (be careful sometimes...)
  
- $\beta_1$ is an _unknown_, constant __slope__

  -   change in average value for $Y$ for each one-unit increase in $X$
  
. . .
  
- $\epsilon_i$ is the __random__ noise

    -   assume __independent, identically distributed__ (_iid_) from a normal distribution 

    -   $\epsilon_i \overset{iid}{\sim}N(0, \sigma^2)$ with constant variance $\sigma^2$



## Simple linear regression estimation

We are estimating the __conditional expection (mean)__ for $Y$:

$$\mathbb{E}[Y_i \mid X_i] = \beta_0 + \beta_1X_i$$

- average value for $Y$ given the value for $X$

. . .

How do we estimate the __best fitted__ line?

. . .

__Ordinary least squares (OLS)__ - by minimizing the __residual sum of squares (RSS)__

$$\text{RSS} \left(\beta_{0}, \beta_{1}\right)=\sum_{i=1}^{n}\left[Y_{i}-\left(\beta_{0}+\beta_{1} X_{i}\right)\right]^{2}=\sum_{i=1}^{n}\left(Y_{i}-\beta_{0}-\beta_{1} X_{i}\right)^{2}$$

. . .

$$\widehat{\beta}_{1}=\frac{\sum_{i=1}^{n}\left(X_{i}-\bar{X}\right)\left(Y_{i}-\bar{Y}\right)}{\sum_{i=1}^{n}\left(X_{i}-\bar{X}\right)^{2}} \quad \text{ and } \quad \widehat{\beta}_{0}=\bar{Y}-\widehat{\beta}_{1} \bar{X}$$

where $\displaystyle \bar{X} = \frac{1}{n}\sum_{i=1}^n X_i$ and $\displaystyle \bar{Y} = \frac{1}{n}\sum_{i=1}^n Y_i$

## Visual explanation

* Imagine a 2D scatterplot with different data points

* Linear regression aims to find a "best fit" line that minimizes the vertical distance between the line and the data points

  * The vertical distance is known as a residual (difference between the actual y-value of a data point and the predicted y-value on the regression line)

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 8
set.seed(2)
d <- tibble(x = rnorm(50, mean = 100, sd = 20)) |>
  mutate(y =  0.8 * x + rnorm(50, mean = 0, sd = 10))
m <- lm(y ~ x, data = d)
d <- d |> 
  mutate(yhat = fitted(m))

basic_scatter <- d |> 
  ggplot(aes(x, y)) +
  geom_point(size = 4) +
  coord_equal()
basic_scatter
```

:::

::: {.column width="50%" style="text-align: left;"}

```{r}
#| echo: false
#| fig-height: 8
basic_scatter +
  geom_segment(aes(xend = x, yend = yhat), color = "red") +
  geom_smooth(method = "lm", color = "red", se = FALSE)
```

:::
:::

# Linear regression in `R`

## Data: Capital Bikeshare Ridership

Daily count of rental bikes between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

```{r load-data, warning = FALSE, message = FALSE}
library(tidyverse)
theme_set(theme_light())

bikes <- read_csv("https://raw.githubusercontent.com/36-SURE/2025/refs/heads/main/data/bikes.csv")
glimpse(bikes)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```


## Modeling bike ridership

::: columns
::: {.column width="50%" style="text-align: left;"}

* To best serve its members, Capital Bikeshare must understand the demand for its service

* Modeling & predicting bikeshare `rides`

```{r bike-rides, echo = TRUE, eval = FALSE}
bikes |>
  ggplot(aes(x = rides)) +
  geom_histogram(color = "black", fill = "gray")
```


:::

::: {.column width="50%" style="text-align: left;"}

```{r ref.label = 'bike-rides', echo = FALSE}

```

:::
:::

## Relationship between rides and feels-like temperature

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r scatter, eval = FALSE}
rides_temp <- bikes |>
  ggplot(aes(x = temp_feel, y = rides)) +
  geom_point(size = 3, alpha = 0.5)
rides_temp
```

We fit linear regression models using `lm()`, formula is input as: `response ~ predictor`

```{r rides-reg}
simple_lm <- lm(rides ~ temp_feel, data = bikes) 
```

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label = 'scatter', echo = FALSE}

```

:::
:::

## Model summary using  `summary()`

::: columns
::: {.column width="45%" style="text-align: left;"}

```{r}
summary(simple_lm)
```

:::

::: {.column width="55%" style="text-align: left;"}

How do we read the summary output?

Model: $\ \small \text{Rides} = \beta_0 + \beta_1 \ \text{Temperature} + \epsilon$

From output: $\ \small \hat{\beta}_0 = -2179.27$ and $\small \hat{\beta}_1 = 81.88$

OLS line: $\ \ \small \widehat{\text{Rides}} = -2179.27 + 81.88 \ \text{Temperature}$

Interpretation of $\hat \beta_1$: estimated change in y for a one unit change in x

In context: For every additional $^{\circ}$ F in feels-like temperature, the expected ridership increases by about 82 rides.

:::
:::

## Inference: hypothesis testing

::: columns
::: {.column width="45%" style="text-align: left;"}

```{r}
summary(simple_lm)
```

:::

::: {.column width="55%" style="text-align: left;"}

Estimates of uncertainty for model coefficients via __standard errors__ $$\quad \widehat{SE}(\hat{\beta}_0) = 358.63 \quad \text{and} \quad \widehat{SE}(\hat{\beta}_1) = 5.12$$

We're testing whether there is a linear relationship between the predictor and the response $$\quad H_0: \beta_1 = 0 \quad \text{vs} \quad H_1: \beta_1 \neq 0$$

Test statistic: $$t = \frac{\hat{\beta}_1}{\widehat{SE}(\hat{\beta}_1)} = \frac{81.88}{5.12} = 15.99$$

:::
:::

## Inference: hypothesis testing

::: columns
::: {.column width="45%" style="text-align: left;"}

```{r}
summary(simple_lm)
```

:::

::: {.column width="55%" style="text-align: left;"}

Recall that we're testing $$\quad H_0: \beta_1 = 0 \quad \text{vs} \quad H_1: \beta_1 \neq 0$$

* $p$-value (i.e., `Pr(>|t|)`): probability of observing a value as extreme as the observed test statistic (under the null hypothesis)
  
* $p$-value $< \alpha = 0.05$ (conventional threshold): __sufficient evidence to reject the null hypothesis that the coefficient is zero__

* In context: There is a statistically significant association between temperature and number of bikeshare rides, $\small \hat \beta_1 = 81.88, t = 15.99,p$-value $<0.001$.

:::
:::
  
## Model summary using the [`broom`](https://cran.r-project.org/web/packages/broom/vignettes/broom.html) package

The 3 `broom` functions (Note: the output is always a tibble)

*   `tidy()`: coefficients table in a tidy format

*   `glance()`: produces summary metrics of a model fit

*   `augment()`: adds/"augments" columns to the original data (e.g., adding predictions)

. . .

```{r}
library(broom)
tidy(simple_lm)
glance(simple_lm)
```

## Confidence interval for model coefficients


```{r}
tidy(simple_lm, conf.int = TRUE, conf.level = 0.95)
```

In-context interpretation (for `temp_feel`)

* For every additional $^{\circ}$ F in feels-like temperature, the expected ridership increases by about 82 rides (95% CI [71.8, 91.9])

. . . 

* We're 95% confident that every additional degree Fahrenheit in temperature is associated with an increase of about 72 to 92 rides in expected ridership.


## Coefficient of determination

* In linear regression, the square of the correlation coefficient happens to be exactly the coefficient of
determination

```{r example-cor}
cor(bikes$temp_feel, bikes$rides)
cor(bikes$temp_feel, bikes$rides) ^ 2
```

* $R^2$ estimates the __proportion of the variance__ of $Y$ explained by $X$ 

* More generally: variance of model predictions / variance of $Y$

```{r general-rsquared}
var(predict(simple_lm)) / var(bikes$rides) 
```

* Interpretation: 33.9% of the variability in the number of bikeshare rides can be explained by the linear relationship between
ridership and feels-like temperature (the remaining 66.1% of the variation is due to other factors and/or noise)

## Generating predictions

We can use the `predict()` or `fitted()` function to get the fitted values from the regression model

```{r init-predict}
head(predict(simple_lm))
```

```{r}
head(fitted(simple_lm))
```

## Plot observed values against predictions

Useful diagnostic (for __any type of model__, not just linear regression!) 

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r plot-pred-obs, eval = FALSE}
bikes |>
  mutate(pred = predict(simple_lm)) |> 
  ggplot(aes(x = pred, y = rides)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_abline(slope = 1, intercept = 0, 
              linetype = "dashed",
              color = "red",
              linewidth = 2)
```

- "Perfect" model will follow __diagonal__

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label='plot-pred-obs', echo = FALSE, fig.height=6,fig.width=6,fig.align="center"}

```

:::
:::



## Plot observed values against predictions

Augment the data with model output using the [`broom` package](https://cran.r-project.org/web/packages/broom/vignettes/broom.html)

`augment()` adds various columns from model fit we can use in plotting for model diagnostics

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r plot-pred-obs-broom, eval = FALSE}
bikes <- simple_lm |> 
  augment(bikes) 
bikes |>
  ggplot(aes(x = .fitted, y = rides)) + 
  geom_point(alpha = 0.5, size = 3) +
  geom_abline(slope = 1, intercept = 0, color = "red", 
              linetype = "dashed", linewidth = 2)
```

:::

::: {.column width="50%" style="text-align: left;"}
```{r ref.label='plot-pred-obs-broom', echo = FALSE, fig.height=6,fig.width=6,fig.align="center"}

```

:::
:::

## Assessing assumptions of linear regression

Simple linear regression assumes $Y_i \overset{iid}{\sim} N(\beta_0 + \beta_1 X_i, \sigma^2)$

- If this is true, then $Y_i - \hat{Y}_i \overset{iid}{\sim} N(0, \sigma^2)$

. . .

Plot residuals against $\hat{Y}_i$, __residuals vs fit__ plot

- Used to assess linearity, any divergence from mean 0

- Used to assess equal variance, i.e., if $\sigma^2$ is homogenous across predictions/fits $\hat{Y}_i$

. . .

More difficult to assess the independence and fixed $X$ assumptions

- Make these assumptions based on subject-matter knowledge



## Plot residuals against predicted values

::: columns
::: {.column width="50%" style="text-align: left;"}

- Residuals = observed - predicted

- Interpretation of residuals in context?

- Conditional on the predicted values, the __residuals should have a mean of zero__

```{r plot-resid-fit-broom, eval = FALSE}
bikes |>
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point(alpha = 0.5, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "red", linewidth = 2) +
  # plot the residual mean
  geom_smooth(se = FALSE)
```

- Residuals __should NOT display any pattern__

- Two things to look for:

  -   Any trend around horizontal reference line?

  -   Equal vertical spread?

:::


::: {.column width="50%" style="text-align: left;"}

```{r ref.label='plot-resid-fit-broom', echo = FALSE, fig.height=6,fig.width=6,fig.align="center"}

```

:::
:::

## Linear regression with one categorical predictor

* Model `rides` by the `season` (4 levels: `spring`, `summer`, `fall`, `winter`)

```{r}
# table(bikes$season)
```

* We are changing the intercept of our regression line by including a categorical variable

* Notice we only get 3 coefficients for the season besides the intercept $$\widehat{\text{Rides}} = 4051 - 265 \ I_{\texttt{spring}} + 543 \ I_{\texttt{summer}} - 1686 \ I_{\texttt{winter}}$$

```{r}
season_lm <- lm(rides ~ season, data = bikes)
tidy(season_lm, conf.int = TRUE)
```

## Linear regression with one categorical predictor

$$\widehat{\text{Rides}} = 4050.7 - 264.7 \ I_{\texttt{spring}} + 543.5 \ I_{\texttt{summer}} - 1686.3 \ I_{\texttt{winter}}$$

* By default, `R` turns the categorical variables of $m$ levels (e.g. 4 seasons in this dataset) into $m-1$ indicator/dummy variables for different categories relative to a baseline/reference level

  * Indicator/dummy variables: binary variables (1: that level, 0: not that level) 

  * `R` uses alphabetical order to determine the reference level (`fall` season in this case)
  
## Linear regression with one categorical predictor

$$\widehat{\text{Rides}} = 4050.7 - 264.7 \ I_{\texttt{spring}} + 543.5 \ I_{\texttt{summer}} - 1686.3 \ I_{\texttt{winter}}$$

* The intercept term gives us the (baseline) average y for the reference level
  
  * In context, the expected daily bikeshare rides during fall season is 4051 rides (95% CI [3841, 4260])

* The values for the coefficient estimates indicate the expected change in the response variable relative to the reference level

  * For example, the expected bikeshare rides during winter is 1686 rides lower (95% CI [-1985, -1388]) than during fall (does this make sense?)
  
## Comparing observed values and predictions

* Similar to before, we can plot the actual and predicted ridership

* To reiterate, all we're doing is changing the intercept of our regression line by including a categorical variable

::: columns
::: {.column width="30%" style="text-align: left;"}

```{r}
#| eval: false
bikes |>
  mutate(season_pred = predict(season_lm)) |> 
  ggplot(aes(x = rides, y = season_pred)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap(~ season, ncol = 4) +
  labs(x = "Actual number of rides", 
       y = "Predicted number of rides")
```

:::


::: {.column width="70%" style="text-align: left;"}

```{r}
#| echo: false
bikes |>
  mutate(season_pred = predict(season_lm)) |> 
  ggplot(aes(x = rides, y = season_pred)) +
  geom_point(size = 1, alpha = 0.5) +
  facet_wrap(~ season, ncol = 4) +
  labs(x = "Actual number of rides", 
       y = "Predicted number of rides")
```

:::

:::

## Interactions

* An interaction allows one predictor’s association with the outcome to depend on values of another predictor

. . .

* Example: interaction between a categorical variable (`season`) and a quantitative variable (`temp_feel`)

  * When to include interaction? When we believe relationships between a quantitative variable and the response differ across categories (of a categorical variable)

  * 4 different lines for each `season` in modeling relationship between `rides` and `temp_feel`

  * This means those regression lines have different intercepts as well as different slopes
  
. . .  
  
* For more on interactions, see these [notes](https://www.medicine.mcgill.ca/epidemiology/joseph/courses/EPIB-621/interaction.pdf)
  
## Visualizing interactions

```{r}
bikes |> 
  ggplot(aes(x = temp_feel, y = rides, color = season)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 2) +
  ggthemes::scale_color_colorblind()
```

## Interpreting interactions

$$
\begin{aligned}
{\text{Rides}} &= \beta_0 + \beta_1 \ \text{Temp} + \beta_2 \ I_{\texttt{spring}} + \beta_3 \ I_{\texttt{summer}} + \beta_4 \ I_{\texttt{winter}} \\
&+ \beta_5 \{ \text{Temp } \times I_{\texttt{spring}} \} +  \beta_6 \{ \text{Temp } \times I_{\texttt{summer}} \} +  \beta_7 \{ \text{Temp } \times I_{\texttt{winter}} \} + \epsilon \\
\end{aligned}
$$

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r}
#| attr-output: "style='font-size: 0.55em;'"
interaction_lm <- lm(rides ~ temp_feel * season, 
                     data = bikes)
tidy(interaction_lm)
```

:::

::: {.column width="50%" style="text-align: left;"}

* $\beta_0$ and $\beta_1$ give the slope and intercept for the regression line (describing the relationship between ridership & temperature) for fall season (the reference level)

* $\beta_2$ measures the difference in the intercepts between spring and fall seasons 

* $\beta_5$ shows how much the slopes change as we move from the regression line for fall (the reference level) to the line for spring

* and so on...

:::
:::


## Multiple linear regression 

We can include as many variables as we want (assuming $n > p$)

$$Y=\beta_{0}+\beta_{1} X_{1}+\beta_{2} X_{2}+\cdots+\beta_{p} X_{p}+\epsilon$$

OLS estimates in matrix notation ($\boldsymbol{X}$ is a $n \times p$ matrix):
  
$$\hat{\boldsymbol{\beta}} = (\boldsymbol{X} ^T \boldsymbol{X})^{-1}\boldsymbol{X}^T\boldsymbol{Y}$$

. . .

Can just add more variables to the formula in `R`

```{r}
multiple_lm <- lm(rides ~ temp_feel + humidity, data = bikes)
```

- Use adjusted $R^2$  when including multiple variables $\displaystyle 1 - \frac{(1 - R^2)(n - 1)}{(n - p - 1)}$

  - Adjusts for the number of variables in the model $p$  
  
  - Adding more variables __will always increase__ the model's (unadjusted) $R^2$
  
READ THIS: [$R^2$ myths](https://www.stat.cmu.edu/~cshalizi/TALR/TALR.pdf#page=181) 

[Get your $R^2$ out of here](https://www.tiktok.com/@fatgreggy/video/7186499956058950958)


## What about the normality assumption?

* Model: $Y=\beta_{0}+\beta_{1} X_{1}+\beta_{2} X_{2}+\cdots+\beta_{p} X_{p}+\epsilon$

- $\epsilon_i$ is the __random__ noise: assume __independent and identically distributed__ (_iid_) from a Normal distribution $\epsilon_i \overset{iid}{\sim}N(0, \sigma^2)$ with constant variance $\sigma^2$

. . .

* __OLS doesn't care about this assumption__, it's just estimating coefficients!

. . .

* In order to perform inference, __we need to impose additional assumptions__

* By assuming $\epsilon_i \overset{iid}{\sim}N(0, \sigma^2)$, what we really mean is $Y \overset{iid}{\sim}N(\beta_{0}+\beta_{1} X_{1}+\beta_{2} X_{2}+\cdots+\beta_{p} X_{p}, \sigma^2)$

. . .

* So we're estimating the mean $\mu$ of this conditional distribution, but what about $\sigma^2$?
  
. . .

* [Unbiased estimate](https://bradleyboehmke.github.io/HOML/linear-regression.html#simple-linear-regression) $\displaystyle \hat{\sigma}^2 = \frac{\text{RSS}}{n - (p + 1)}$

* __Degrees of freedom__:  $n - (p + 1)$, data supplies us with $n$ "degrees of freedom" and we used up $p + 1$


## Regression confidence intervals vs prediction intervals

Do we want to predict the mean response for a particular value $x^*$ of the explanatory variable or do we want to predict the response for an individual case?

Example: 

*   if we would like to know the average ridership for **all** days with feels-like temperature of 70 degrees Fahrenheit, then we would use an interval for the **mean response**

*   if we want an interval to contain the ridership of **a particular** day with feels-like temperature of 70 degrees Fahrenheit, then we need the interval for **a single prediction**


## Regression confidence intervals

::: columns
::: {.column width="50%" style="text-align: left;"}

- `geom_smooth()` displays __confidence intervals__ for the regression line

```{r conf-int, eval = FALSE}
# predict(simple_lm, interval = "confidence")
lm_plot <- bikes |>
  ggplot(aes(x = temp_feel, y = rides)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
lm_plot
```

:::


::: {.column width="50%" style="text-align: left;"}

```{r, ref.label='conf-int', echo = FALSE, fig.height=5, fig.width=6,fig.align="center"}

```

:::
:::


---

## Regression confidence intervals

*   Interval estimate for the MEAN response at a given observation (i.e. for the predicted AVERAGE $\hat{\beta}_0 + \hat{\beta}_1 x^*$)

*   Based on standard errors for the estimated regression line at $x^*$ $$SE_{\text{line}}\left(x^{*}\right)=\hat{\sigma} \cdot \sqrt{\frac{1}{n}+\frac{\left(x^{*}-\bar{x}\right)^{2}}{\sum_{i=1}^{n}\left(x_{i}-\bar{x}\right)^{2}}}$$

*   Variation only comes from uncertainty in the parameter 

## Regression prediction intervals

*   interval estimate for an INDIVIDUAL response at a given (new) observation 

*   add the variance $\sigma^2$ of a __single predicted value__ $$SE_{\text{pred}}\left(x^{*}\right)=\hat{\sigma} \cdot \sqrt{1 + \frac{1}{n}+\frac{\left(x^{*}-\bar{x}\right)^{2}}{\sum_{i=1}^{n}\left(x_{i}-\bar{x}\right)^{2}}}$$

*   Variation comes from uncertainty in parameter estimates and error variance 

## Confidence intervals versus prediction intervals

Generate 95% intervals

::: columns
::: {.column width="50%" style="text-align: left;"}

```{r pred-interval, eval = FALSE}
# predict(simple_lm, interval = "prediction")
lm_plot +
  geom_ribbon(
    data = augment(simple_lm, interval = "prediction"),
    aes(ymin = .lower, ymax = .upper),
    color = "red", fill = NA
  )
```

*   The standard error for predicting an individual response will always be larger than for predicting a mean response

*   Prediction intervals will always be wider than confidence intervals

:::

::: {.column width="50%" style="text-align: left;"}

```{r, ref.label='pred-interval', echo = FALSE,  fig.height=6,fig.width=6,fig.align="center"}

```

:::
:::
